---
title: "speciation_summary.RMD"
output: html_document
date: "2023-06-21"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

## Speciation events within a range
The speciation log file contains records of all speciation events that happened within a range. 
It records the species of the range (ancestor) and all descendants. 
It also notes the nodes and branch lengths (in time scale) that separate the ancestor and descendant species.
```{r}
library(coda)
log <-  read.table("morph_at_start/penguins_inf_morph_at_start_combined.speciation.log",
                   header=TRUE, sep="\t")
# burnIn <- 0.1
# log <- tail(log, -round(nrow(log)*burnIn)) # take out 10% burnin
log<-log[,1:(ncol(log)-1)]
columnsWithValue <- colnames(log)[colSums(log != "0.0,0.0") != 0]
# we don't need "Sample" column
columnsWithValue <- columnsWithValue[2:length(columnsWithValue)]
log <- log[, columnsWithValue]
```

```{r}
n <- length(columnsWithValue)
d <- data.frame(from=character(n), to=character(n), prob=numeric(n), length_median=numeric(n),
                length_hpd_l=numeric(n), length_hpd_h=numeric(n), nodes_median=numeric(n),
                nodes_hpd_l=numeric(n), nodes_hpd_h=numeric(n))
# Split each string by ":"
split_strings <- strsplit(columnsWithValue, "\\.")
# Extract the first and second elements
d$from <- sapply(split_strings, function(x) x[1])
d$to <- sapply(split_strings, function(x) x[2])

for (i in 1:n){
  split_strings <- strsplit(log[,columnsWithValue[i]], ",")
  length <- sapply(split_strings, function(x) as.double(x[1]))
  nodes <- sapply(split_strings, function(x) as.double(x[2]))
  
  d$prob[i] <- length(length)
  idx <- which(length != 0)
  length <- length[idx] # no length means no speciation between the two species recorded.
  nodes <- nodes[idx] # 0 intermediate nodes are legit
  d$prob[i] <- length(length)/d$prob[i]
  
  if (length(idx)>1){
    d$length_median[i] <- median(length)
    length_hpd <- HPDinterval(as.mcmc(length))
    d$length_hpd_l[i] <- length_hpd[1]
    d$length_hpd_h[i] <- length_hpd[2]
    
    d$nodes_median[i] <- median(nodes)
    nodes_hpd <- HPDinterval(as.mcmc(nodes))
    d$nodes_hpd_l[i] <- nodes_hpd[1]
    d$nodes_hpd_h[i] <- nodes_hpd[2]
  }
}
```

Sort for highest posterior probability of divergence.
```{r}
d <- d[order(d$prob,decreasing = TRUE),]
head(d, 30)
```


## Comaparison of the clades obtained by SA and sRanges 
The next log file is obtained by running a SRangesAndSACladeSetComparator tool on trees obtained by Sampled ancestor
and sRanges packages. For all clades that were found, the posterior probability of the clade, its height and HPD interval bounds for the height are recorded.
```{r}
# this is already with bunin
log <-  read.table("figures/clade_comp.txt", header=FALSE, sep=" ")
colnames(log) <- c("Species", "PosteriorProbFBD", "PosteriorProbSRanges",
                   "HeightFBD", "HeightFBD_HPD_l", "HeightFBD_HPD_h",
                   "HeightSRFBD", "HeightSRFBD_HPD_l", "HeightSRFBD_HPD_h")


```


You can obtain the clades with very low probabilities in both sets
```{r}
thr <- 0.2 
log_thr <- log[which(log$PosteriorProbFBD>thr & log$PosteriorProbSRanges>thr),]
head(log_thr)
```

You can obtain the clades that have posterior support difference greater than some threshold:
```{r}
library(stringr)
# thr <- 0.15 
# log_thr <- log[which(abs(log$PosteriorProbFBD-log$PosteriorProbSRanges)>thr),]
log_thr$probDiff <- abs(log_thr$PosteriorProbFBD-log_thr$PosteriorProbSRanges)
log_thr$nSpecies <- count.fields(textConnection(log_thr$Species), sep = ",")
b<-lapply(log_thr$Species, function(x) str_remove_all(x, "[{}]"))
log_thr$cladeName <- lapply(b, function(x) str_flatten(unique(sapply(unlist(strsplit(x, ',')),
                                                                     function(y) str_split(y, "_")[[1]])[1,]), "--"))

library(ggplot2)
library(ggExtra)
require(cowplot)
require(forcats)
l <- rep("", nrow(log_thr))
l[which(log_thr$HeightFBD-log_thr$HeightSRFBD>0)]<-"\u2194"
l<-l[order(log_thr$probDiff)]
p<-ggplot(log_thr, aes(x = HeightFBD_HPD_l, y = fct_reorder(Species, probDiff))) +
  geom_segment(aes(xend = HeightFBD_HPD_h, yend = fct_reorder(Species, probDiff), colour = "fbd"), alpha=0.4,  linewidth = 1, ) +
  geom_point(aes(x = HeightFBD, shape="fbd"), size = 1) +
  geom_segment(aes(x=HeightSRFBD_HPD_l, xend = HeightSRFBD_HPD_h, yend = fct_reorder(Species, probDiff), colour = "srfbd"), linewidth = 1, alpha=0.4) +
  geom_point(aes(x = HeightSRFBD, shape="srfbd"), size = 1) +
  scale_color_manual(values = c(fbd = "blue", srfbd = "red"),
                     labels = c(fbd = "FBD", srfbd = "SRFBD"),
                     name = "")+
  scale_shape_manual(values = c(fbd = 4, srfbd = 3),
                     labels = c(fbd = "FBD", srfbd = "SRFBD"),
                     name = "")+
  theme_bw() + xlab("Age (Ma)") + ylab("Clades") +
  theme(plot.margin=unit(c(0.2, -0.2, 0.2, 0.2),"cm"),
        axis.ticks.y = element_blank(), legend.direction="horizontal", legend.position = c(0.5, 0.07))  + scale_y_discrete(
    name = "",
    labels = l, position="right"
  )
require(grDevices)
p2 <- ggplot(log_thr, aes(x = 0, y = fct_reorder(Species, probDiff))) +
  geom_segment(aes(xend = PosteriorProbFBD, yend = fct_reorder(Species, probDiff)), linewidth = 1, alpha=0.4, colour = "blue")+
  geom_segment(aes(x=0, xend = PosteriorProbSRanges, yend = fct_reorder(Species, probDiff)), linewidth = 1, alpha=0.4, colour = "red")+
  theme_bw() + xlab("Posterior probability") +
  theme(legend.position = "none",axis.text.y = element_blank(),
        axis.title.y=element_blank(), axis.ticks.y = element_blank(), 
        plot.margin=unit(c(0.2, 0.2, 0.2, -0.2),"cm"))
# ggMarginal(data=log_thr, x="HeightFBD_HPD_l", y="PosteriorProbFBD", margins="y")
p3<-plot_grid(p, p2, ncol=2, rel_widths = c(3,1))
p3
pdf("figures/clade_prob_thr_both_0.2.pdf")
print(p3)
dev.off()
ggsave2("figures/clade_prob_thr_both_0.2.pdf", p3, width=5, height=5, device=cairo_pdf, family="Arial Unicode MS")
log_thr <- apply(log_thr,2,as.character)
for (i in 2:10){
  log_thr[,i]<-round(as.numeric(log_thr[,i]),2)
}
log_thr <- log_thr[order(log_thr[,10], decreasing = TRUE),]
log_thr[,4] <- paste0(as.character(log_thr[,4]), " [", as.character(log_thr[,5]),
                      ", ", as.character(log_thr[,6]), "]")
log_thr[,7] <- paste0(as.character(log_thr[,7]), " [", as.character(log_thr[,8]),
                      ", ", as.character(log_thr[,9]), "]")
log_thr <- log_thr[,-c(5,6,8,9)]
write.csv2(log_thr, "figures/clade_prob_thr_both_0.2.csv")
head(log_thr)
log_thr_1 <- log[which(log$PosteriorProbFBD>=0.01 & log$PosteriorProbSRanges>=0.01),]
mm <- mean(abs(log_thr_1$HeightFBD-log_thr_1$HeightSRFBD))
write.csv2(mm, "figures/mean_height_diff_thr_0.01.csv")
```

Or filter the log to only keep clades associated with probability more than some threshold
```{r}
thr <- 0.1 
log_thr_both <- log[which(log$PosteriorProbFBD>thr | log$PosteriorProbSRanges>thr),]
log_thr_both$nSpecies <- count.fields(textConnection(log_thr_both$Species), sep = ",")
head(log_thr_both)
```

```{r}
thr <- 0.1 
log_thr_h <- log[which(log$PosteriorProbSA>thr | log$PosteriorProbSRanges>thr),]
log_thr_h <- log_thr_h[which(abs(log_thr_h$HeightFBD-log_thr_h$HeightSRFBD)>thr),]
log_thr_h$heightDiff <- abs(log_thr_h$HeightFBD-log_thr_h$HeightSRFBD)
head(log_thr_h)
```

Here we try to find the most probable clades of less than 15 species and involving Delphinornis gracilis
```{r}
Delphinornis_gracilis<- log_thr_both[grep("Delphinornis_gracilis", log_thr_both$Species),]
Delphinornis_gracilis <- Delphinornis_gracilis[which(Delphinornis_gracilis$nSpecies<15),]
print(Delphinornis_gracilis)
```

Here, we analyse clades involving Madrynornis mirandus
```{r}
Madrynornis_mirandus<- log_thr_both[grep("Madrynornis_mirandus", log_thr_both$Species),]
Madrynornis_mirandus <- Madrynornis_mirandus[which(Madrynornis_mirandus$PosteriorProbFBD>thr | Madrynornis_mirandus$PosteriorProbSRanges>0.1),]
Madrynornis_mirandus$probDiff <- abs(Madrynornis_mirandus$PosteriorProbFBD-Madrynornis_mirandus$PosteriorProbSRanges)
Madrynornis_mirandus_thr <- Madrynornis_mirandus[which(Madrynornis_mirandus$probDiff>0.1),]
b<-lapply(Madrynornis_mirandus_thr$Species, function(x) str_remove_all(x, "[{}]"))
Madrynornis_mirandus_thr$cladeName <- lapply(b, function(x) str_flatten(unique(sapply(unlist(strsplit(x, ',')),
                                                                     function(y) str_split(y, "_")[[1]])[1,]), "--"))
print(Madrynornis_mirandus_thr)
```


We next look at the Eudyptes clade.
```{r}
eudyptes_str<- "Eudyptes_chrysocome,Eudyptes_chrysolophus,Eudyptes_filholi,Eudyptes_moseleyi,Eudyptes_pachyrhynchus,Eudyptes_robustus,Eudyptes_schlegeli,Eudyptes_sclateri"
eudyptes<- log_thr_both[grep(eudyptes_str, log_thr_both$Species),]
print(eudyptes)
```
```{r}
Marplesornis_novaezealandiae<- log_thr_both[grep("Marplesornis_novaezealandiae", log_thr_both$Species),]
Marplesornis_novaezealandiae <- Marplesornis_novaezealandiae[which(Marplesornis_novaezealandiae$PosteriorProbSA>thr | Marplesornis_novaezealandiae$PosteriorProbSRanges>0.1),]
Marplesornis_novaezealandiae$probDiff <- abs(Marplesornis_novaezealandiae$PosteriorProbSA-Marplesornis_novaezealandiae$PosteriorProbSRanges)
Marplesornis_novaezealandiae_thr <- Marplesornis_novaezealandiae[which(Marplesornis_novaezealandiae$probDiff>0.1),]
print(Marplesornis_novaezealandiae_thr)
```


# ```{r}
# # this is already with bunin
# log_extant <-  read.table("figures2/extant_clade_comp.txt", header=FALSE, sep=" ")
# colnames(log_extant) <- c("Species", "PosteriorProbSA", "PosteriorProbSRanges",
#                    "HeightSA", "HeightSA_HPD_l", "HeightSA_HPD_h",
#                    "HeightSRanges", "HeightSRanges_HPD_l", "HeightSRanges_HPD_h")
# log_extant$nSpecies <- count.fields(textConnection(log_extant$Species), sep = ",")
# log_extant$probDiff <- abs(log_extant$PosteriorProbSA-log_extant$PosteriorProbSRanges)
# thr <- 0.4 
# log_extant_thr <- log_extant[which(log_extant$PosteriorProbSA>thr | log_extant$PosteriorProbSRanges>thr),]
# log_extant_thr <- log_extant[which(log_extant$probDiff>0.1),]
# 
# Madrynornis_mirandus <- log_extant[grep("Madrynornis_mirandus", log_extant$Species, ignore.case = FALSE),]
# Madrynornis_mirandus <- Madrynornis_mirandus[which(Madrynornis_mirandus$PosteriorProbSA>thr | Madrynornis_mirandus$PosteriorProbSRanges>0.1),]
# 
# 
# Megadyptes_antipodes <- log_extant[grep("Megadyptes_antipodes", log_extant$Species, ignore.case = FALSE),]
# Megadyptes_antipodes <- Megadyptes_antipodes[which(Megadyptes_antipodes$PosteriorProbSA>thr | Megadyptes_antipodes$PosteriorProbSRanges>0.1),]
# ```